---
title: "Lab 4"
author: "Anders Lie"
date: "3/19/2021"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(rmarkdown)
def <- readxl::read_excel('cyclonesFootball2020.xlsx', sheet='Defensive')
ofn <- readxl::read_excel('cyclonesFootball2020.xlsx', sheet='Offensive')
bio <- readxl::read_excel('cyclonesFootball2020.xlsx', sheet='Biography')
```

# Part One: Data Cleaning

## 1
```{r}
def <- def %>% 
  mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent))
ofn <- ofn %>% 
  mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent))
bio <- bio %>% 
  mutate(Name=as.factor(Name))
```

Here we simply use as.factor to turn the character columns into factor columns.

## 2
```{r}
def <- def %>%
  mutate(across(Tackles_Solo:Pass_PB, as.numeric))
ofn <- ofn %>%
  mutate(across(Receiving_REC:Passing_INT, as.numeric))
```

Using across and as.numeric, we easily convert all player statistics to numeric.

## 3
```{r}
bio <- bio %>%
  separate(Height, into=c('HeightFeet', 'HeightInches'), sep='-') %>%
  mutate(across(c(HeightFeet, HeightInches), as.numeric)) %>%
  mutate(Height=12*HeightFeet + HeightInches) %>%
  select(-HeightFeet, -HeightInches)
```

I convert height to be inches only by separating the string and then
recombining using inches = 12 times feet plus extra inches.

# Part Two: Tidying Data

## 1

```{r}
defClean <- def %>%
  pivot_longer(Tackles_Solo:Pass_PB, names_to="stat")

offClean <- ofn %>%
  pivot_longer(Receiving_REC:Passing_INT, names_to="stat")

str(defClean)
str(offClean)
```

Here we see that defClean and offClean contain the data in a long format.

## 2

```{r}
offClean %>% ggplot(aes(x=value)) + geom_histogram() + facet_wrap(~stat) +
  xlab('Value of Statistic') + ylab('Number of Players with Value')
```

We see that the number of Passing_INT, Passing_TD, Receiving_REC, Receiving_TD, and Rushing_TD is relatively
low across players, whereas statistics like Passing_YDS, Receiving_YDS, Rushing_ATT, and
Rushing_YDS are more broadly distributed.

## 3

```{r}
offClean %>%
  filter(Opponent_Opponent %in% c('Oregon', 'Oklahoma')) %>%
  filter(stat=='Receiving_YDS') %>%
  distinct(Name, Opponent_Opponent, .keep_all=TRUE) %>%
  pivot_wider(names_from=Opponent_Opponent, values_from=value) %>%
  ggplot(aes(x=Oregon, y=Oklahoma)) + geom_point() +
  xlab('Receiving Yards against Oregon') +
  ylab('Receiving Yards against Oklahoma') 
```

It seems as though in general, ISU did better against Oklahoma, since
there were a few players who did around 40 yards against Oklahoma as opposed
to around 20 against Oregon, as well as a couple outliers who had around 60 and 80
against Oklahoma nd only around 50 against Oregon.


## 4
```{r}
bio1 <- bio %>%
  separate(Hometown, into=c('City', 'State'), sep=', ')
head(bio1)
```

Here we can see that the hometown has been separated into City and State columns. 

## 5

```{r}
bio1 %>%
  ggplot(aes(x=fct_infreq(State))) + geom_bar() + coord_flip() +
  ylab('Number of Players') + xlab('State')
```

Using our new state column, we can create the above barchart showing how many ISU
players are from each state. Obviously, most ISU players come from Iowa.

# Part Three: Joining Data Frames

## 1

```{r}
# States of Defensive players
def_states <- def %>% left_join(bio1) %>% distinct(Name, .keep_all=TRUE)

def_states %>%
  ggplot(aes(x=fct_infreq(State))) + geom_bar() + coord_flip() +
  ylab('Number of Defensive Players') + xlab('State')

# States of Offensive players
ofn_states <- ofn %>% left_join(bio1) %>% distinct(Name, .keep_all=TRUE)

ofn_states %>%
  ggplot(aes(x=fct_infreq(State))) + geom_bar() + coord_flip() +
  ylab('Number of Offensive Players') + xlab('State')
```

In the barchart, we can see that offensive players come from a wider range of states, whereas
defense players come from only a handful. Interestingly, a plurality of players on the
offensive come from Florida, whereas players on the defensive come mostly from Iowa.

# 2

```{r}
purdy_def <- def %>% filter(Name == 'Purdy, Brock')

not_zero <- function(x){return(x != 0)}

purdy_def %>% mutate(across(Tackles_Solo:Pass_PB, not_zero)) %>% paged_table
```


We can see there is only one nonzero statistic for Purdy on defense, so there
is not enough data to try and determine his defensive contribution.
Therefore we will focus on his offensive contributions.

We will compare the performance of Purdy to the rest of the team by showing
the all the teams statistics, with Purdy's highlighted in red:

```{r}
cols <- c("Purdy, Brock"="red", 
          "Akers, Landen"="grey", 
          "Brock, Jirehl"="grey", 
          "Hutchinson, Xavier"="grey", 
          "Nwangwu, Kene"="grey", 
          "Scates, Joe"="grey", 
          "Soehner, Dylan"="grey", 
          "Hall, Breece"="grey", 
          "Dekkers, Hunter"="grey", 
          "Kolar, Charlie"="grey", 
          "Allen, Chase"="grey", 
          "Shaw Jr., Sean"="grey", 
          "Milton, Tarique"="grey", 
          "Bitter, Aidan"="grey", 
          "Jackson, Daniel"="grey", 
          "Dean, Easton"="grey")

offClean %>%
  group_by(Name, stat) %>%
  summarize(avg_val=mean(value, na.rm=TRUE)) %>%
  ggplot(aes(x=Name, weight=avg_val, fill=Name)) + geom_bar() + 
  facet_wrap(~stat) + coord_flip() +
  scale_fill_manual(values=cols)
```

We can see that, compared to the rest of the team, Purdy has far more passing yards.
Other than that, he has no particularly large statistics. I believe this is because
he is the quarterback.

## 3

Note that we will only compare ISU players who have records in both 2019 and 2020,
otherwise of course we cannot see how much they improved.

```{r}
def2019 <- readxl::read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
ofn2019 <- readxl::read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
bio2019 <- readxl::read_excel('cyclonesFootball2019.xlsx', sheet='Biography')

def2019 <- def2019 %>% 
  mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent))
ofn2019 <- ofn2019 %>% 
  mutate(Name=as.factor(Name), Opponent_Opponent=as.factor(Opponent_Opponent))
bio2019 <- bio2019 %>% 
  mutate(Name=as.factor(Name))

def2019 <- def2019 %>%
  mutate(across(Tackles_Solo:Pass_PB, as.numeric))
ofn2019 <- ofn2019 %>%
  mutate(across(Rushing_ATT:Passing_INT, as.numeric))

def2019_long <- def2019 %>%
  pivot_longer(Tackles_Solo:Pass_PB, names_to="stat")

ofn2019_long <- ofn2019 %>%
  pivot_longer(Rushing_ATT:Passing_INT, names_to="stat")

def2019avg <- def2019_long %>%
  filter(Name %in% defClean$Name) %>%
  group_by(Name, stat) %>%
  summarize(avg_val=mean(value))

def2020avg <- defClean %>%
  filter(Name %in% def2019_long$Name) %>%
  group_by(Name, stat) %>%
  summarize(avg_val=mean(value))

cmp_def <- def2020avg
# Avg amount improved
cmp_def$avg_val = def2020avg$avg_val - def2019avg$avg_val
cmp_def %>%
  ggplot(aes(x=Name, weight=avg_val)) + geom_bar() + facet_wrap(~stat) + coord_flip()
```

Here we can see that the player Jake Hummel had a lot of big increases in these statistics,
particularly improving the most in assisted tackles, and improving the second most in solo tackles.

JaQuan Bailey would have been another candidate, however his assistent tackles decreased significantly.

So I would say Jake Hummel improved the most as a defensive player.

```{r}
off2019avg <- ofn2019_long %>%
  filter(Name %in% defClean$Name) %>%
  group_by(Name, stat) %>%
  summarize(avg_val=mean(value))

off2020avg <- offClean %>%
  filter(Name %in% def2019_long$Name) %>%
  group_by(Name, stat) %>%
  summarize(avg_val=mean(value))

cmp_off <- off2020avg
cmp_off$avg_val = off2020avg$avg_val - off2019avg$avg_val
cmp_off %>%
  ggplot(aes(x=Name, weight=avg_val)) + geom_bar() + facet_wrap(~stat) + coord_flip()
```

For offensive players who played in 2019 and 2020, it is quite clear that Breece Hall
is the player who improved the most, having the greatest improvements over any other player
in every single offensive statistic besides receiving yards.





